Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2017-02-02 aosd13/paperBody.tex
Revise related work for AOSD13

\chapter{Collections as a case study}
\label{sec:caseStudy}
In this chapter we discuss collections as a \emph{critical} case
study~\citep{flyvbjerg06five} for deep DSL embedding.

@@

\section{Deep Embeddings in Scala}
\label{sec:rwdsl}
Technically, our implementation of \LoS\ is a deep embedding of a part of the Scala collections API~\citep{odersky2009fighting}.
Deep embeddings were pionereed by \citet{Leijen99DSEC} and \citet{elliott03compiling}. \pg{This makes no sense for the thesis.}
The technical
details of the embedding are not the main topic of this chapter; we are using some of the
Scala techniques presented by \citet{rompf2010lightweight} for using implicits and for
adding infix operators to a type. Similar to \citet{rompf2010lightweight}, we also 
use the Scala compiler on-the-fly. A plausible alternative backend for \LoS\ would
have been to use Delite~\citep{Rompf11BBlocks}, a framework for
building highly efficient DSLs in Scala.
Using this framework, in concurrent work, \citet{Rompf13} also optimize
collection queries; while their work allows for imperative programs, they do
not support embedding arbitrary libraries in an automated way. On the other
hand, they can reuse support for automatic parallelization and multiple platforms present in Delite.
\citet{Ackermann12} present Jet, which also optimizes collection queries but
targets MapReduce-style computations in a distributed environment.
Moreover, both works do not apply typical database optimizations such as
indexing or filter hoisting.

We regard the Scala collections API~\citep{odersky2009fighting} as a shallowly embedded query DSL\@. Query operators immediately perform collection operations when called, so that it is not possible to optimize queries before execution. In addition to these eager query operators, the Scala collections API also provides \emph{views} to create lazy collections.
Views are somewhat similar to {\LoS} in that they reify query operators as data structures and interpret them later.
However, views are not used for automatic query optimization, but for explicitly changing the evaluation order of collection processing. Unfortunately, views are not suited as a basis for the implementation of {\LoS} because they only reify the outermost pipeline of collection operators, whereas nested collection operators as well as other Scala code in queries, such as filter predicates or \code{map} and \code{flatMap} arguments, are only shallowly embedded.
Deep embedding of the whole query is necessary for many optimizations, as discussed in Sec.~\ref{sec:solution}.

\smartParagraph{Code Querying}
In our evaluation we explore the usage of \LoS\ to express queries on code and re-implement a subset of the FindBugs~\citep{DBLP:journals/sigplan/HovemeyerP04} analyses. There are various other specialized code query languages such as 
CodeQuest~\citep{Hajiyev06CodeQuest} or D-CUBED~\citep{Wegrzynowicz:2009:GBU:1639950.1640032}. 
Since these are special-purpose query languages that are not embedded into a host language, they are not directly comparable to our approach.

>>>>

\chapter{Collection query embedding as a case study}
\label{sec:caseStudy}
In this chapter we discuss collections as a \emph{critical} case
study~\citep{flyvbjerg06five} for deep DSL embedding.

@@

\section{Deep Embeddings in Scala}
\label{sec:rwdsl}
Technically, our implementation of \LoS\ is a deep embedding of a part of the Scala collections API~\citep{odersky2009fighting}.
Deep embeddings were pionereed by \citet{Leijen99DSEC} and \citet{elliott03compiling} in Haskell for other applications.

We regard the Scala collections API~\citep{odersky2009fighting} as a shallowly
embedded query DSL\@. Query operators are \emph{eager}, that is they immediately
perform collection operations when called, so that it is not possible to
optimize queries before execution.

Scala collections also provide \emph{views}, which are closer to {\LoS}. Unlike
standard query operators, they create \emph{lazy} collections. Like {\LoS},
views reify query operators as data structures and interpret them later. Unlike
{\LoS}, views are not used for automatic query optimization, but for explicitly
changing the evaluation order of collection processing.
%
Moreover, they cannot be reused by {\LoS} as they reify too little: Views embed
deeply only the outermost pipeline of collection operators, while they embed
shallowly nested collection operators and other Scala code in queries, such as
arguments of \code{filter}, \code{map} and \code{flatMap}.
%
Deep embedding of the whole query is necessary for many optimizations, as
discussed in Sec.~\ref{sec:solution}.

\subsection{LMS}
Our deep embedding is inspired by some of the Scala techniques presented by
Lightweight Modular Staging (LMS)~\citep{rompf2010lightweight} for using
implicits and for adding infix operators to a type. Like
\citet{rompf2010lightweight}, we also generate and compile Scala code on-the-fly
reusing the Scala compiler. A plausible alternative backend for \LoS\ would have
been to use LMS and Delite~\citep{Rompf11BBlocks}, a framework for building
highly efficient DSLs in Scala.

An alternative to \LoS\ based on LMS and Delite, named OptiQL, was indeed built
and presented in concurrent~\citep{Rompf13} and subsequent work~\citep{Arvind13}.
Like \LoS, OptiQL enables writing and optimizing collection queries in Scala.
On the minus side, OptiQL supports fewer collections types (\texttt{ArrayList} and \texttt{HashMap}).

On the plus side, OptiQL supports queries containing effects and can reuse support
for automatic parallelization and multiple platforms present in Delite.
While LMS allows embedding effectful queries, it is unclear how many of the
implemented optimizations keep being sound on such queries, and how many of
those have been extended to such queries.

OptiQL achieves impressive speedups by fusing collection operations and
transforming (or \emph{lowering}) high-level bulk operators to highly optimized
imperative programs using \texttt{while} loops. This gains significant
advantages because it can avoid boxing, intermediate allocations, and because
inlining heuristics in the Hotspot JIT compiler have never been tuned to bulk
operators in functional programs.\footnote{See discussion by Cliff Click at \url{http://www.azulsystems.com/blog/cliff/2011-04-04-fixing-the-inlining-problem}.}\pg{Maybe elaborate and cite Cliff Click's blog post, turn it into a citation?}
We did not investigate such lowerings. It is unclear how
well these optimizations would extend to other collections which intrinsically
carry further overhead.
\pg{Resume and connect with \cref{sec:optim-vs-inc}.}
\pg{Cite Stream fusion to completeness~\citep{Kiselyov2017stream}!}

Those works did not support embedding arbitrary libraries in an automated or
semi-automated way; this was only addressed later in
Forge~\citep{Sujeeth13Forge} and Yin-Yang \citep{Jovanovic2014}.

\citet{Ackermann12} present Jet, which also optimizes collection queries but
targets MapReduce-style computations in a distributed environment.
Like OptiQL, Jet does not apply typical database optimizations such as
indexing or filter hoisting.

\section{Code Querying}
In our evaluation we explore the usage of \LoS\ to express queries on code and re-implement a subset of the FindBugs~\citep{DBLP:journals/sigplan/HovemeyerP04} analyses. There are various other specialized code query languages such as 
CodeQuest~\citep{Hajiyev06CodeQuest} or D-CUBED~\citep{Wegrzynowicz:2009:GBU:1639950.1640032}. 
Since these are special-purpose query languages that are not embedded into a host language, they are not directly comparable to our approach.
